---
title: Effects of Weather Events on Population Health and Local Economies
author: davedev44 @ github
output: html_document
---

## Summary

This document contains analysis of major weather events as they affected
population health (injuries and fatalities) as well as their economic
consequences (estimated monetary value of property and crop damage). Through
investigation of data provided by the NOAA between 1996 and 2011, it was
determined that tornados had the greatest effect on population health, while
flooding had the greatest economic consequences.

## Collaboration Note

I drew heavily upon the "Project 2 Tips" thread in the discussion forums for
completion of this report. In particular, the ideas for cleaning up the
data categories (and how to do that with approximate string matching), using 
only events from 1996-2011 (and the source for that), and using 
inflation-adjusted dollar figures. In addition, as in other projects, I drew
upon various internet sources for little things (how to do this or that in R).

## Data Processing

The NOAA dataset is obtained as a bzfile from a designated source on the web,
obtainable as of the data of the production of this document. The data
were then loaded.

```{r getloaddata, echo=TRUE}

# URL for the data of interest
NOAA_URL <- 
"https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"

filename <- "StormData.bz2"

# Obtain the file
download.file(NOAA_URL, filename)

# Read the file. Note that read.csv can read a compressed file!
stormdata <- read.csv(filename)
```

The dataset is a list of major weather events between 1950 and 2011. In order 
to address the issues of interest (effects of different types of storm events 
on population health and local economies), we reduce the data set to the 
columns of interest. Let's take a look at the names of the columns in the 
dataset.

```{r, echo=TRUE}
names(stormdata)
```

Based on the above names, we would want to extract the EVTYPE column (event
type), FATALITIES and INJURIES (as a measure of effect on population health),
and PROPDMG/PROPDMGEXP and CROPDMG/CROPDMGEXP as measures of economic effects
(property damage and crop damage for a given event). PROPDMG and CROPDMG
are numeric values, paired with their respective "EXP" variables as a multiplier
(for instance, if PROPDMG is 5, and PROPDMGEXP is "k", the monetary damage
would be 5000). We'll also retain the date column; its cleanup will be
discussed later.

So we extract the above-listed columns.

```{r, echo=TRUE}
stormdataSub <- stormdata[,c("EVTYPE", 
                             "FATALITIES", 
                             "INJURIES", 
                             "PROPDMG",
                             "PROPDMGEXP",
                             "CROPDMG", 
                             "CROPDMGEXP",
                             "BGN_DATE")]
```


Next, we'll do some data cleaning. First, we'll clean up the dates. According
to related documentation 
(http://www.ncdc.noaa.gov/stormevents/details.jsp?type=eventtype) tracking
of all events did not start until 1996, and as such, events before 1996
will be overweighted toward only those types of events that were tracked.
As such, we restrict our analysis to 1996-2011.

```{r, echo=TRUE}

# Clean up the date column by removing the time portion and converting to a date.
stormdataSub$BGN_DATE <- as.Date(gsub("(\\S+) \\S+", "\\1", stormdataSub$BGN_DATE, perl=TRUE), "%m/%d/%Y")

# Remove all records before 1996
effectiveDate <- as.Date("01/01/1996", "%m/%d/%Y")
stormdataSub <- stormdataSub[stormdataSub$BGN_DATE >= effectiveDate,]

# Add an additional column with just the year, for later inflation calculations
stormdataSub$year <- format(stormdataSub$BGN_DATE, "%Y")
```

Now we set about cleaning up columns of interest. First, the event types.
The database documentation (https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf) 
provides 48 "official" categories; in the
EVTYPE column there are many typos, errors, and the like; the cleaning
effort for this column will be the process of converting all of the values
in the EVTYPE column to one of the official categories. We do so by trimming whitespace,
converting values to upper case, eliminating values (events) for which
the mapping does not make sense, using an approximate string match as a
first cut for establishing the intended mapping, and finally, manually
inspecting the first cut mapping and editing as necessary.

The mapping will by no means be perfect, but it is reproducible with the
following code, and is a best attempt.
```{r, echo=TRUE}

# The list of 48 "official" event types
officialEventTypes = c ("ASTRONOMICAL LOW TIDE", 
               "AVALANCHE",
               "BLIZZARD", 
               "COASTAL FLOOD",
               "COLD/WIND CHILL",
               "DEBRIS FLOW", 
               "DENSE FOG",
               "DENSE SMOKE",
               "DROUGHT",
               "DUST DEVIL",
               "DUST STORM",
               "EXCESSIVE HEAT",
               "EXTREME COLD/WIND CHILL",
               "FLASH FLOOD",
               "FLOOD",
               "FREEZING FOG",
               "FROST/FREEZE",
               "FUNNEL CLOUD",
               "HAIL", 
               "HEAT", 
               "HEAVY RAIN",
               "HEAVY SNOW",
               "HIGH SURF",
               "HIGH WIND", 
               "HURRICANE/TYPHOON",
               "ICE STORM",
               "LAKESHORE FLOOD",
               "LAKE-EFFECT SNOW",
               "LIGHTNING", 
               "MARINE HAIL", 
               "MARINE HIGH WIND", 
               "MARINE STRONG WIND", 
               "MARINE THUNDERSTORM WIND", 
               "RIP CURRENT", 
               "SEICHE", 
               "SLEET", 
               "STORM TIDE",
               "STRONG WIND", 
               "THUNDERSTORM WIND", 
               "TORNADO", 
               "TROPICAL DEPRESSION", 
               "TROPICAL STORM", 
               "TSUNAMI", 
               "VOLCANIC ASH", 
               "WATERSPOUT", 
               "WILDFIRE", 
               "WINTER STORM", 
               "WINTER WEATHER")

# Convert to ALL CAPS and trim whitespace
library(stringr)
stormdataSub$EVTYPEADJ <- str_trim(toupper(stormdataSub$EVTYPE))

# Immediately remove events containing SUMMARY, MONTHLY, NONE, OTHER, NORTHERN LIGHTS
# as they are not trying to communicate an event type
stormdataSub <- stormdataSub[!grepl("SUMMARY", stormdataSub$EVTYPEADJ),]
stormdataSub <- stormdataSub[!grepl("MONTHLY", stormdataSub$EVTYPEADJ),]
stormdataSub <- stormdataSub[!grepl("NONE", stormdataSub$EVTYPEADJ),]
stormdataSub <- stormdataSub[!grepl("OTHER", stormdataSub$EVTYPEADJ),]
stormdataSub <- stormdataSub[!grepl("NORTHERN LIGHTS", stormdataSub$EVTYPEADJ),]

# stringdist gives a warning about version number that is not relevant
suppressWarnings(library(stringdist))
library(plyr)

# Compute frequency of event types. We don't care about the counts (freq column),
# just the unique list of event types (adjusted so far as above)
rawEventFreq <- count(stormdataSub, 'EVTYPEADJ')
rawEventFreq$freq <- NULL

# Match each of those to the official event types with an approximate string match
# Picked maxDist 15 as it seems to strike a balance between finding a match
# for mostly everything, and making mostly correct matches. It does not matter too
# much as it will be inspected by hand and adjusted anyway.
officialEventMatch <- amatch(rawEventFreq$EVTYPEADJ, officialEventTypes, maxDist=15)

# Attach the result of the match as a column to the event frequency
rawEventFreq$matchedEventType <- officialEventTypes[officialEventMatch]

# Make manual corrections - a mixture of regexp and exact matches
rawEventFreq$matchedEventType[grep("TSTM", rawEventFreq$EVTYPEADJ)] <- "THUNDERSTORM WIND"
rawEventFreq$matchedEventType[which(rawEventFreq$EVTYPEADJ=="ABNORMAL WARMTH")] <- "HEAT"
rawEventFreq$matchedEventType[which(rawEventFreq$EVTYPEADJ=="ABNORMALLY DRY")] <- "DROUGHT"
rawEventFreq$matchedEventType[which(rawEventFreq$EVTYPEADJ=="ABNORMALLY WET")] <- "HEAVY RAIN"
rawEventFreq$matchedEventType[which(rawEventFreq$EVTYPEADJ=="ASTRONOMICAL HIGH TIDE")] <- "STORM TIDE"
rawEventFreq$matchedEventType[which(rawEventFreq$EVTYPEADJ=="BEACH EROSION")] <- "HIGH SURF"
rawEventFreq$matchedEventType[which(rawEventFreq$EVTYPEADJ=="BITTER WIND CHILL TEMPERATURES")] <- "COLD/WIND CHILL"
rawEventFreq$matchedEventType[which(rawEventFreq$EVTYPEADJ=="BLACK ICE")] <- "ICE STORM"
rawEventFreq$matchedEventType[which(rawEventFreq$EVTYPEADJ=="BLOWING DUST")] <- "DUST STORM"
rawEventFreq$matchedEventType[grep("^COLD", rawEventFreq$EVTYPEADJ)] <- "COLD/WIND CHILL"
rawEventFreq$matchedEventType[which(rawEventFreq$EVTYPEADJ=="COOL SPELL")] <- "COLD/WIND CHILL"
rawEventFreq$matchedEventType[which(rawEventFreq$EVTYPEADJ=="DAM BREAK")] <- "FLOOD"
rawEventFreq$matchedEventType[which(rawEventFreq$EVTYPEADJ=="DOWNBURST")] <- "THUNDERSTORM WIND"
rawEventFreq$matchedEventType[which(rawEventFreq$EVTYPEADJ=="DRIEST MONTH")] <- "DROUGHT"
rawEventFreq$matchedEventType[which(rawEventFreq$EVTYPEADJ=="DRIFTING SNOW")] <- "HEAVY SNOW"
rawEventFreq$matchedEventType[which(rawEventFreq$EVTYPEADJ=="DROWNING")] <- "FLOOD"
rawEventFreq$matchedEventType[which(rawEventFreq$EVTYPEADJ=="DRY")] <- "DROUGHT"
rawEventFreq$matchedEventType[grep("MICROBURST", rawEventFreq$EVTYPEADJ)] <- "TORNADO"
rawEventFreq$matchedEventType[which(rawEventFreq$EVTYPEADJ=="DRY WEATHER")] <- "DROUGHT"
rawEventFreq$matchedEventType[which(rawEventFreq$EVTYPEADJ=="EARLY FROST")] <- "COLD/WIND CHILL"
rawEventFreq$matchedEventType[which(rawEventFreq$EVTYPEADJ=="EXCESSIVE COLD")] <- "EXTREME COLD/WIND CHILL"
rawEventFreq$matchedEventType[grep("^EXCESSIVE RAIN", rawEventFreq$EVTYPEADJ)] <- "HEAVY RAIN"
rawEventFreq$matchedEventType[which(rawEventFreq$EVTYPEADJ=="EXCESSIVE SNOW")] <- "HEAVY SNOW"
rawEventFreq$matchedEventType[which(rawEventFreq$EVTYPEADJ=="EXCESSIVELY DRY")] <- "DROUGHT"
rawEventFreq$matchedEventType[which(rawEventFreq$EVTYPEADJ=="EXTENDED COLD")] <- "COLD/WIND CHILL"
rawEventFreq$matchedEventType[which(rawEventFreq$EVTYPEADJ=="EXTREME COLD")] <- "EXTREME COLD/WIND CHILL"
rawEventFreq$matchedEventType[which(rawEventFreq$EVTYPEADJ=="EXTREME WINDCHILL TEMPERATURES")] <- "EXTREME COLD/WIND CHILL"
rawEventFreq$matchedEventType[which(rawEventFreq$EVTYPEADJ=="EXTREMELY WET")] <- "HEAVY RAIN"
rawEventFreq$matchedEventType[which(rawEventFreq$EVTYPEADJ=="FALLING SNOW/ICE")] <- "ICE STORM"
rawEventFreq$matchedEventType[which(rawEventFreq$EVTYPEADJ=="FIRST FROST")] <- "FROST/FREEZE"
rawEventFreq$matchedEventType[which(rawEventFreq$EVTYPEADJ=="FLOOD/STRONG WIND")] <- "FLOOD"
rawEventFreq$matchedEventType[which(rawEventFreq$EVTYPEADJ=="FOG")] <- "DENSE FOG"
rawEventFreq$matchedEventType[which(rawEventFreq$EVTYPEADJ=="FREEZE")] <- "FROST/FREEZE"
rawEventFreq$matchedEventType[grep("^FREEZING", rawEventFreq$EVTYPEADJ)] <- "SLEET"
rawEventFreq$matchedEventType[which(rawEventFreq$EVTYPEADJ=="FROST")] <- "FROST/FREEZE"
rawEventFreq$matchedEventType[which(rawEventFreq$EVTYPEADJ=="GLAZE")] <- "ICE STORM"
rawEventFreq$matchedEventType[grep("THUNDERSTORM WIND", rawEventFreq$EVTYPEADJ)] <- "THUNDERSTORM WIND"
rawEventFreq$matchedEventType[which(rawEventFreq$EVTYPEADJ=="GUSTY WIND/HAIL")] <- "HAIL"
rawEventFreq$matchedEventType[which(rawEventFreq$EVTYPEADJ=="GUSTY WIND/HVY RAIN")] <- "HEAVY RAIN"
rawEventFreq$matchedEventType[which(rawEventFreq$EVTYPEADJ=="GUSTY WIND/RAIN")] <- "HEAVY RAIN"
rawEventFreq$matchedEventType[grep("^HAIL", rawEventFreq$EVTYPEADJ)] <- "HAIL"
rawEventFreq$matchedEventType[which(rawEventFreq$EVTYPEADJ=="HEAVY SEAS")] <- "HIGH SURF"
rawEventFreq$matchedEventType[which(rawEventFreq$EVTYPEADJ=="HEAVY SURF")] <- "HIGH SURF"
rawEventFreq$matchedEventType[which(rawEventFreq$EVTYPEADJ=="HEAVY SURF AND WIND")] <- "HIGH SURF"
rawEventFreq$matchedEventType[which(rawEventFreq$EVTYPEADJ=="HIGH WATER")] <- "FLOOD"
rawEventFreq$matchedEventType[grep("^HOT", rawEventFreq$EVTYPEADJ)] <- "HEAT"
rawEventFreq$matchedEventType[which(rawEventFreq$EVTYPEADJ=="HURRICANE")] <- "HURRICANE/TYPHOON"
rawEventFreq$matchedEventType[which(rawEventFreq$EVTYPEADJ=="HYPERTHERMIA/EXPOSURE")] <- "EXCESSIVE HEAT"
rawEventFreq$matchedEventType[which(rawEventFreq$EVTYPEADJ=="HYPOTHERMIA/EXPOSURE")] <- "EXTREME COLD/WIND CHILL"
rawEventFreq$matchedEventType[which(rawEventFreq$EVTYPEADJ=="ICE")] <- "ICE STORM"
rawEventFreq$matchedEventType[which(rawEventFreq$EVTYPEADJ=="ICE FOG")] <- "FREEZING FOG"
rawEventFreq$matchedEventType[grep("^ICE JAM", rawEventFreq$EVTYPEADJ)] <- "FLOOD"
rawEventFreq$matchedEventType[grep("^LANDSLIDE", rawEventFreq$EVTYPEADJ)] <- "DEBRIS FLOW"
rawEventFreq$matchedEventType[which(rawEventFreq$EVTYPEADJ=="LANDSLUMP")] <- "DEBRIS FLOW"
rawEventFreq$matchedEventType[which(rawEventFreq$EVTYPEADJ=="LANDSPOUT")] <- "TORNADO"
rawEventFreq$matchedEventType[which(rawEventFreq$EVTYPEADJ=="LATE SEASON HAIL")] <- "HAIL"
rawEventFreq$matchedEventType[grep("^LATE SEASON SNOW", rawEventFreq$EVTYPEADJ)] <- "HEAVY SNOW"
rawEventFreq$matchedEventType[which(rawEventFreq$EVTYPEADJ=="LATE-SEASON SNOWFALL")] <- "HEAVY SNOW"
rawEventFreq$matchedEventType[which(rawEventFreq$EVTYPEADJ=="LIGHT FREEZING RAIN")] <- "SLEET"
rawEventFreq$matchedEventType[grep("^LIGHT SNOW", rawEventFreq$EVTYPEADJ)] <- "HEAVY SNOW"
rawEventFreq$matchedEventType[which(rawEventFreq$EVTYPEADJ=="MARINE ACCIDENT")] <- "MARINE STRONG WIND"
rawEventFreq$matchedEventType[which(rawEventFreq$EVTYPEADJ=="MARINE THUNDERSTORM WIND")] <- "MARINE THUNDERSTORM WIND"
rawEventFreq$matchedEventType[which(rawEventFreq$EVTYPEADJ=="MARINE TSTM WIND")] <- "MARINE THUNDERSTORM WIND"
rawEventFreq$matchedEventType[which(rawEventFreq$EVTYPEADJ=="METRO STORM, MAY 26")] <- "THUNDERSTORM WIND"
rawEventFreq$matchedEventType[which(rawEventFreq$EVTYPEADJ=="MILD AND DRY PATTERN")] <- "DROUGHT"
rawEventFreq$matchedEventType[which(rawEventFreq$EVTYPEADJ=="MINOR FLOODING")] <- "FLOOD"
rawEventFreq$matchedEventType[grep("^MIXED PRECIP", rawEventFreq$EVTYPEADJ)] <- "SLEET"
rawEventFreq$matchedEventType[which(rawEventFreq$EVTYPEADJ=="MUD SLIDE")] <- "DEBRIS FLOW"
rawEventFreq$matchedEventType[grep("MUDSLIDE", rawEventFreq$EVTYPEADJ)] <- "DEBRIS FLOW"
rawEventFreq$matchedEventType[which(rawEventFreq$EVTYPEADJ=="NON SEVERE HAIL")] <- "HAIL"
rawEventFreq$matchedEventType[which(rawEventFreq$EVTYPEADJ=="NON TSTM WIND")] <- "HIGH WIND"
rawEventFreq$matchedEventType[which(rawEventFreq$EVTYPEADJ=="NON-SEVERE WIND DAMAGE")] <- "HIGH WIND"
rawEventFreq$matchedEventType[which(rawEventFreq$EVTYPEADJ=="NON-TSTM WIND")] <- "HIGH WIND"
rawEventFreq$matchedEventType[which(rawEventFreq$EVTYPEADJ=="PATCHY ICE")] <- "ICE STORM"
rawEventFreq$matchedEventType[which(rawEventFreq$EVTYPEADJ=="PROLONG COLD")] <- "EXTREME COLD/WIND CHILL"
rawEventFreq$matchedEventType[which(rawEventFreq$EVTYPEADJ=="PROLONG WARMTH")] <- "EXCESSIVE HEAT"
rawEventFreq$matchedEventType[grep("^RAIN", rawEventFreq$EVTYPEADJ)] <- "HEAVY RAIN"
rawEventFreq$matchedEventType[grep("RECORD*COLD", rawEventFreq$EVTYPEADJ)] <- "EXTREME COLD/WIND CHILL"
rawEventFreq$matchedEventType[which(rawEventFreq$EVTYPEADJ=="RECORD COOL")] <- "EXTREME COLD/WIND CHILL"
rawEventFreq$matchedEventType[grep("^RECORD DRY", rawEventFreq$EVTYPEADJ)] <- "DROUGHT"
rawEventFreq$matchedEventType[which(rawEventFreq$EVTYPEADJ=="RECORD HEAT")] <- "EXCESSIVE HEAT"
rawEventFreq$matchedEventType[which(rawEventFreq$EVTYPEADJ=="RECORD HIGH")] <- "EXCESSIVE HEAT"
rawEventFreq$matchedEventType[which(rawEventFreq$EVTYPEADJ=="RECORD LOW RAINFALL")] <- "DROUGHT"
rawEventFreq$matchedEventType[grep("^RECORD WARM", rawEventFreq$EVTYPEADJ)] <- "EXCESSIVE HEAT"
rawEventFreq$matchedEventType[which(rawEventFreq$EVTYPEADJ=="RECORD WINTER SNOW")] <- "HEAVY SNOW"
rawEventFreq$matchedEventType[grep("^RED FLAG", rawEventFreq$EVTYPEADJ)] <- "WILDFIRE"
rawEventFreq$matchedEventType[which(rawEventFreq$EVTYPEADJ=="REMNANTS OF FLOYD")] <- "HURRICANE/TYPHOON"
rawEventFreq$matchedEventType[grep("^RIVER FLOOD", rawEventFreq$EVTYPEADJ)] <-"FLOOD"
rawEventFreq$matchedEventType[which(rawEventFreq$EVTYPEADJ=="ROCK SLIDE")] <- "DEBRIS FLOW"
rawEventFreq$matchedEventType[which(rawEventFreq$EVTYPEADJ=="ROGUE WAVE")] <- "HIGH SURF"
rawEventFreq$matchedEventType[which(rawEventFreq$EVTYPEADJ=="ROUGH SEAS")] <- "MARINE STRONG WIND"
rawEventFreq$matchedEventType[which(rawEventFreq$EVTYPEADJ=="SAHARAN DUST")] <- "DUST STORM"
rawEventFreq$matchedEventType[grep("^SEVERE THUNDERSTORM", rawEventFreq$EVTYPEADJ)] <- "THUNDERSTORM WIND"
rawEventFreq$matchedEventType[grep("^SLEET", rawEventFreq$EVTYPEADJ)] <- "SLEET"
rawEventFreq$matchedEventType[which(rawEventFreq$EVTYPEADJ=="SMALL HAIL")] <- "HAIL"
rawEventFreq$matchedEventType[which(rawEventFreq$EVTYPEADJ=="SML STREAM FLD")] <- "FLOOD"
rawEventFreq$matchedEventType[which(rawEventFreq$EVTYPEADJ=="SMOKE")] <- "WILDFIRE"
rawEventFreq$matchedEventType[grep("^SNOW", rawEventFreq$EVTYPEADJ)] <- "HEAVY SNOW"
rawEventFreq$matchedEventType[which(rawEventFreq$EVTYPEADJ=="SNOWMELT FLOODING")] <- "FLOOD"
rawEventFreq$matchedEventType[which(rawEventFreq$EVTYPEADJ=="STREET FLOODING")] <- "FLOOD"
rawEventFreq$matchedEventType[which(rawEventFreq$EVTYPEADJ=="TEMPERATURE RECORD")] <- "EXCESSIVE HEAT"
rawEventFreq$matchedEventType[which(rawEventFreq$EVTYPEADJ=="THUNDERSNOW SHOWER")] <- "HEAVY SNOW"
rawEventFreq$matchedEventType[which(rawEventFreq$EVTYPEADJ=="TORRENTIAL RAINFALL")] <- "HEAVY RAIN"
rawEventFreq$matchedEventType[which(rawEventFreq$EVTYPEADJ=="TYPHOON")] <- "HURRICANE/TYPHOON"
rawEventFreq$matchedEventType[grep("^UNSEASON*COLD", rawEventFreq$EVTYPEADJ)] <- "EXTREME COLD/WIND CHILL"
rawEventFreq$matchedEventType[grep("^UNSEASON*COOL", rawEventFreq$EVTYPEADJ)] <- "EXTREME COLD/WIND CHILL"
rawEventFreq$matchedEventType[which(rawEventFreq$EVTYPEADJ=="UNSEASONABLY COOL & WET")] <- "EXTREME COLD/WIND CHILL"
rawEventFreq$matchedEventType[which(rawEventFreq$EVTYPEADJ=="UNSEASONABLY DRY")] <- "DROUGHT"
rawEventFreq$matchedEventType[which(rawEventFreq$EVTYPEADJ=="UNSEASONABLY HOT")] <- "HEAT"
rawEventFreq$matchedEventType[grep("^UNSEASONABLY WARM", rawEventFreq$EVTYPEADJ)] <- "HEAT"
rawEventFreq$matchedEventType[which(rawEventFreq$EVTYPEADJ=="UNSEASONABLY WET")] <- "HEAVY RAIN"
rawEventFreq$matchedEventType[which(rawEventFreq$EVTYPEADJ=="UNSEASONAL LOW TEMP")] <- "EXTREME COLD/WIND CHILL"
rawEventFreq$matchedEventType[which(rawEventFreq$EVTYPEADJ=="UNSEASONAL RAIN")] <- "HEAVY RAIN"
rawEventFreq$matchedEventType[which(rawEventFreq$EVTYPEADJ=="UNUSUAL WARMTH")] <- "HEAT"
rawEventFreq$matchedEventType[which(rawEventFreq$EVTYPEADJ=="UNUSUAL/RECORD WARMTH")] <- "HEAT"
rawEventFreq$matchedEventType[which(rawEventFreq$EVTYPEADJ=="UNUSALLY COLD")] <- "EXTREME COLD/WIND CHILL"
rawEventFreq$matchedEventType[which(rawEventFreq$EVTYPEADJ=="UNUSALLY WARM")] <- "HEAT"
rawEventFreq$matchedEventType[grep("URBAN*FLOOD", rawEventFreq$EVTYPEADJ)] <- "FLOOD"
rawEventFreq$matchedEventType[grep("URBAN*FLD", rawEventFreq$EVTYPEADJ)] <- "FLOOD"
rawEventFreq$matchedEventType[which(rawEventFreq$EVTYPEADJ=="VERY DRY")] <- "DROUGHT"
rawEventFreq$matchedEventType[which(rawEventFreq$EVTYPEADJ=="VERY WARM")] <- "HEAT"
rawEventFreq$matchedEventType[which(rawEventFreq$EVTYPEADJ=="VOG")] <- "DENSE FOG"
rawEventFreq$matchedEventType[which(rawEventFreq$EVTYPEADJ=="WARM WEATHER")] <- "HEAT"
rawEventFreq$matchedEventType[which(rawEventFreq$EVTYPEADJ=="WET MICOBURST")] <- "TORNADO"
rawEventFreq$matchedEventType[which(rawEventFreq$EVTYPEADJ=="WET MONTH")] <- "HEAVY RAIN"
rawEventFreq$matchedEventType[which(rawEventFreq$EVTYPEADJ=="WET YEAR")] <- "HEAVY RAIN"
rawEventFreq$matchedEventType[which(rawEventFreq$EVTYPEADJ=="WHIRLWIND")] <- "TORNADO"
rawEventFreq$matchedEventType[grep("^WIND", rawEventFreq$EVTYPEADJ)] <- "HIGH WIND"
rawEventFreq$matchedEventType[which(rawEventFreq$EVTYPEADJ=="WND")] <- "HIGH WIND"

# Now, merge the correspondence of official event type and event types found in the
# original data back into the original data
stormdataSub <- merge(stormdataSub, rawEventFreq)

# Convert resultant "official" matched type to a factor
stormdataSub$matchedEventType <- as.factor(stormdataSub$matchedEventType)
```

Next we look at cleaning up and computing economic effects, as measured
by dollar amounts of damage ot crops and property. This is measured by the
damage-EXP variables; we find a number of different levels, varying case 
("K" and "k"). Since the only meaningful levels are billions ("B"), 
millions ("M"), thousands ("K"), and no multiplier (""), we take steps to 
remove records that contain EXP values other than those, and compute numeric 
damage estimates based on those EXP values.

Begin with property damage:

```{r echo=TRUE}
levels(stormdataSub$PROPDMGEXP)

# Convert to character in order to find empty strings more easily
stormdataSub$PROPDMGEXP <- as.character(stormdataSub$PROPDMGEXP)

# Convert the EXP variable to upper case to standardize
stormdataSub$PROPDMGEXP <- toupper(stormdataSub$PROPDMGEXP)

# Generate a "fake" character value for empty string ("D" for "dollars")
stormdataSub$PROPDMGEXP[nchar(stormdataSub$PROPDMGEXP)==0] <- "D"

# Reduce the dataset to only the values of interest / values that make
# sense
stormdataSub <- stormdataSub[stormdataSub$PROPDMGEXP %in% c("B", "M", "K", "D"),]

# Replace the letter values with corresponding numeric values
stormdataSub$PROPDMGEXP[stormdataSub$PROPDMGEXP=="B"] <- "1000000000"
stormdataSub$PROPDMGEXP[stormdataSub$PROPDMGEXP=="M"] <- "1000000"
stormdataSub$PROPDMGEXP[stormdataSub$PROPDMGEXP=="K"] <- "1000"
stormdataSub$PROPDMGEXP[stormdataSub$PROPDMGEXP=="D"] <- "1"

# Establish as a numeric value
stormdataSub$PROPDMGEXP <- as.numeric(stormdataSub$PROPDMGEXP)

# Compute the product of the numeric PROPDMG with the now numeric
# multiplier PROPDMGEXP
stormdataSub$PROPDMGVAL <- stormdataSub$PROPDMG * stormdataSub$PROPDMGEXP
```

Exactly the same for crop damage:

```{r echo=TRUE}
levels(stormdataSub$CROPDMGEXP)

# Convert to character in order to find empty strings more easily
stormdataSub$CROPDMGEXP <- as.character(stormdataSub$CROPDMGEXP)

# Convert the EXP variable to upper case to standardize
stormdataSub$CROPDMGEXP <- toupper(stormdataSub$CROPDMGEXP)

# Generate a "fake" character value for empty string ("D" for "dollars")
stormdataSub$CROPDMGEXP[nchar(stormdataSub$CROPDMGEXP)==0] <- "D"

# Reduce the dataset to only the values of interest / values that make
# sense
stormdataSub <- stormdataSub[stormdataSub$CROPDMGEXP %in% c("B", "M", "K", "D"),]

# Replace the letter values with corresponding numeric values
stormdataSub$CROPDMGEXP[stormdataSub$CROPDMGEXP=="B"] <- "1000000000"
stormdataSub$CROPDMGEXP[stormdataSub$CROPDMGEXP=="M"] <- "1000000"
stormdataSub$CROPDMGEXP[stormdataSub$CROPDMGEXP=="K"] <- "1000"
stormdataSub$CROPDMGEXP[stormdataSub$CROPDMGEXP=="D"] <- "1"

# Establish as a numeric value
stormdataSub$CROPDMGEXP <- as.numeric(stormdataSub$CROPDMGEXP)

# Compute the product of the numeric CROPDMG with the now numeric
# multiplier CROPDMGEXP
stormdataSub$CROPDMGVAL <- stormdataSub$CROPDMG * stormdataSub$CROPDMGEXP
```

Finally, both for crop and property damage, we want to adjust the dollar
value figures for inflation, so we obtain dollar inflation data from 1996-2011,
based on 2011 dollars.

The data can be obtained manually at the following link: 
http://liberalarts.oregonstate.edu/files/polisci/faculty-research/sahr/inflation-conversion/pdf/cv2011.pdf

We obtain this data, which expresses the value of one 2011 dollar in dollars
for other years. We must invert those figures to determine the value of one
dollar from other years in 2011 dollars; this forms the "inflation multiplier"
for each event in our data set, that will be multiplied by the prior-computed
damage estimates.

```{r getinflationdata, echo=TRUE}
year <- c(1996:2011)

# Data are provided as factors (as in, how many dollars
# in year X are equal to 1 dollar in 2011), so we invert
# to get the inflation multiplier (e.g. 1 dollar in year
# X is equal to how many dollars in 2011)
inflationFactors <- c(0.698,
                      0.714,
                      0.725,
                      0.741,
                      0.766,
                      0.787,
                      0.800,
                      0.818,
                      0.840,
                      0.868,
                      0.896,
                      0.922,
                      0.957,
                      0.954,
                      0.969,
                      1.000)

# Round to 3 decimal places to match raw data
inflationMultiplier <- round(1 / inflationFactors, 3)
inflationData <- cbind(year, inflationMultiplier)

# Merge the inflation data in by year (already in our dataset)
stormdataSub <- merge(stormdataSub, inflationData)

# For each event, compute inflation-adjusted dollar figures based on
# the merged inflation multipliers
stormdataSub$PROPDMGVALINFL <- stormdataSub$PROPDMGVAL * stormdataSub$inflationMultiplier
stormdataSub$CROPDMGVALINFL <- stormdataSub$CROPDMGVAL * stormdataSub$inflationMultiplier
```

## Analysis

The questions of interest were: which types of events had the greatest effect
on population health, and which types of events had the greatest economic
consequences.

The effect on population health can be measured by the combined number
of fatalities and injuries per event type. As such, the events are categorized
by event type, with the sum of fatalities and injuries computed for each
event type.

```{r echo=TRUE}
library(plyr)

healthEffectByEventType <- ddply(stormdataSub, 
                                 .(matchedEventType),
                                 summarize,
                                 healthEffect=sum(FATALITIES+INJURIES))
```

The economic effects can be measured by the combined amounts of property and
crop damage per event type. Once again, we categorize the events by event
type, and sum the inflation-adjusted crop and property damage values as computed 
above.

```{r echo=TRUE}
econEffectByEventType=ddply(stormdataSub, 
                            .(matchedEventType), 
                            summarize, 
                            econEffect=sum(PROPDMGVALINFL+CROPDMGVALINFL))
```

## Results

The above categorizations will allow us to answer the questions of interest.

### Health Effects

First, let's look at a plot of the health effect "values" (combined fatalities
and injuries per event type).

```{r echo=TRUE}
# In order to make the plot look better, skip x-axis (since it is obvious
# from the plot title, and from the content of the axis)

# To ensure the x-axis labels don't get cut off, establish an outer margin
par(oma=c(4,1,0,1))

plot(healthEffectByEventType, 
     ylab="Sum of Fatalities and Injuries", 
     xlab="",
     main="Health Effects (Fatalities & Injuries)\n of Weather Events by Event Type, 1996-2011",
     xaxt="n")

# Custom x-axis with perpendicular labels, and shrunken font size to ensure they fit
axis(side="1", 
     las=2, 
     at=healthEffectByEventType$matchedEventType, 
     labels=levels(healthEffectByEventType$matchedEventType),
     cex.axis=0.5)
```

We can read off this plot that the greatest health effect was observed for
tornados.

### Economic Effects

We can similarly look at the economic effects of each type of weather event.

```{r echo=TRUE}
par(oma=c(4,1,0,1))

plot(econEffectByEventType, 
     ylab="Combined Crop and Property Damage (2011 $)", 
     xlab="",
     main="Economic Effects (Crop & Property Damage)\n of Weather Events by Event Type, 1996-2011",
     xaxt="n")

# Custom x-axis with perpendicular labels, and shrunken font size to ensure they fit
# Also adjust 
axis(side="1", 
     las=2, 
     at=econEffectByEventType$matchedEventType, 
     labels=levels(econEffectByEventType$matchedEventType),
     cex.axis=0.5)
```

We can read off this plot that the greatest economic effect was observed for
flooding.

